\documentclass[conference]{IEEEtran}

% Essential packages
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{url}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{multirow}

% Fix for itemize spacing
\setlist[itemize]{leftmargin=*,itemsep=2pt,topsep=4pt}

\begin{document}

% Title
\title{Physics-Guided Self-Supervised Graph Neural Networks for Power Grid Analysis:\\
Transfer Learning Across Cascade Prediction, Power Flow, and Line Flow}

% Authors - to be filled in
\author{
    \IEEEauthorblockN{Anonymous Authors}
    \IEEEauthorblockA{\textit{For Review}}
}

\maketitle

% Abstract
\begin{abstract}
Modern power grid operations require rapid computational methods for decision support, but traditional solvers face severe bottlenecks. We present a physics-guided self-supervised learning framework for graph neural networks that addresses labeled data scarcity in power system analysis. Our approach embeds electrically-parameterized message passing into the encoder architecture---where edge weights are learned from line admittance features---and develops grid-specific pretext tasks (masked injection and parameter reconstruction) enabling representation learning from unlabeled operational data. Evaluated on the PowerGraph benchmark, our method achieves substantial improvements in low-label regimes: 29.1\% power flow error reduction, 26.4\% line flow error reduction, and 6.8\% F1-score improvement for cascading failure prediction---all at 10\% labeled data availability. On the IEEE 118-bus system under severe class imbalance, self-supervised pretraining dramatically reduces training instability (variance reduction from $\pm$0.243 to $\pm$0.051) while improving mean performance ($\Delta$F1 = +0.61). We achieve 0.93 AUC-ROC explainability fidelity via Integrated Gradients, addressing a recognized gap in interpretable cascade prediction.
\end{abstract}

\begin{IEEEkeywords}
Power systems, graph neural networks, self-supervised learning, cascading failures, power flow, line flow prediction, explainability, physics-informed machine learning
\end{IEEEkeywords}

% Main sections
\input{01_introduction.tex}
\input{02_related_work.tex}
\input{03_problem_formulation.tex}
\input{04_methodology.tex}
\input{05_experimental_setup.tex}
\input{06_results.tex}
\input{07_discussion.tex}
\input{08_conclusion.tex}

% Acknowledgments (optional)
\section*{Acknowledgments}
This work was conducted using the PowerGraph benchmark dataset. We thank the developers of PyTorch Geometric for their open-source graph neural network library.

% Bibliography
\bibliographystyle{IEEEtran}
\bibliography{citations}

\end{document}
